---
title: "Day 1: Introduction to RStudio & the Tidyverse"
author: "Kathleen Hablutzel"
date: "6/8/2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Welcome!!!

#### Getting Started

If you haven't already, start downloading R and R Studio. See the instructions [here](index.html) under "Getting Started."

Today's complete code is linked [here]().

#### Creating an RMarkdown Document

* Write text, embed code chunks
* R is the language (in console and code chunks)
* RStudio is the programming environment
* RMarkdown is a markup language for combining text with code
    + Produce reproducible reports where the code is right there
    + Export to HTML, PDF, MS Word (how I created website)
    
```{r}
summary(cars)
```

See the RMarkdown [reference guide](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf).

## Data Visualizations with ggplot2

We will:
* Make a plot
* Add basic plot elements
* Add colors (colorbrewer)
* Add a theme -> point to documentation
* Extension: copy the master examples (a couple graphs for students to copy, then I'll post the code next week and they can see how close they got)

#### Introducing... the Tidyverse

Plotting with base R is… not super fun. 

```{r}
plot(pressure)
```

Instead, we plot with `ggplot2`, a popular open-source package. `ggplot2` is an implementation of Wilkinson's *Grammar of Graphics*, which means that you combine plot building blocks to create your visualization.

`ggplot2` is part of the Tidyverse, a series of open-source R packages that “share an underlying design philosophy, grammar, and data structures”. (See the [website](https://www.tidyverse.org/) and [GitHub repository](https://github.com/tidyverse/tidyverse).)

The main packages are `ggplot2`, `dplyr`, `tidyr`, `readr`, `purrr`, `tibble`, `stringr`, and `forcats`, but we will mostly focus on **ggplot2** (for data visualization) and **dplyr** (for data wrangling).

Let's load the tidyverse:

```{r}
library(tidyverse)
```

In the bottom right panel, under the Packages tab, the `tidyverse` packages should now be checked. In each file, you need to load packages before you use them.

#### Your First Plot

We are going to work with the built-in `mtcars` data set.

```{r Scatterplot}
# a basic plot
my_plot <- ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# adding color
my_plot <- ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue")

# mapping a variable to color
my_plot <- ggplot(data = mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point()

my_plot
```

###### Available Data Sets

The `data()` function will show you all your available data sets:

```{r}
data()
```

The `iris` data set is a famous data set used in machine learning classes. Let's take a look at the documentation:

```{r}
?iris
```

Now, look at the sources. This data comes from the work of a famous eugenicist, statistician, and geneticist, R.A. Fisher, and was published in a journal on eugenics. Many statisticians are currently pushing to use alternate data sets in their classrooms and [remove Fisher's name](https://www.change.org/p/american-statistical-association-rename-the-fisher-lecture-after-david-blackwell) from the prestigious [Fisher Lectureship](https://en.wikipedia.org/wiki/R._A._Fisher_Lectureship).

Data is more than numbers.

## Visual Cues

* Strongest/weakest cues
* Why stacked bar charts >>>>>>>> pie charts

Ranking visual cues, from position (strongest) to color hue (weakest): docs/day-1_files/figure-html/visual-cues.png (link not  working rn)
Source: Nathan Yau, *Data Points*, 2013