---
title: 'Day 4: Spatial Data'
author: "Kathleen Hablutzel"
date: "7/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
library(tidyverse)
library(nycflights13)
library(sf)
library(leaflet)
```

## Page Under Construction

## Getting Started

Here is the recording of today's lesson (*coming soon*). (This is only accessible within the NCSSM organization.) You can also follow along with the code below.

Today we will:

* Create static maps using `sf` and `ggplot2`
* Create interactive maps using `leaflet`
* Learn version control with GitHub

You'll need to install the `sf` and `leaflet` packages for later in the lesson.

#### Notes Moving Forward

**Schedule:** Next week will be our last week of content. We will cover assorted R programming concepts, such as functions and iteration, and we might need to push our GitHub lesson to next week if we spend too long mapping.

**Resources:** I am compiling a list of further R references and resources, as well as an FAQ from student questions throughout this course.

**Guarantee:** This website will stay on my GitHub as long as my GitHub exists, so this course will be available for years to come. I might *slightly* rearrange content so this website becomes a self-explanatory course, and I will convert the lessons to YouTube videos (with all student information removed) so the lectures will be available as well.

#### Last Week's Challenges

Last week, I challenged you to find assorted information about the relational data in `nycflights13` and fix a couple untidy data sets. The solutions are  long enough that I have moved the solutions to the [bottom of this page](https://krhablutzel.github.io/srip-bootcamp/day-4.html#solutions_for_last_week).

## Static Maps with `sf`



## Interactive Maps with `leaflet`

## GitHub

What is Git?
Terminal or GitHub Desktop
straight from RStudio

## Challenges for Next Week

Let's apply what we just learned! For next week, try to

As always, please Slack me throughout the week with any questions! I'll reveal the code for these challenges next week.

#### Solutions For Last Week

Finally, here are the solutions to last week's challenges:

```{r Challenge-Solutions}
# What is the average departure delay for each origin airport?
dep_delay <- flights %>%
  group_by(origin) %>%
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE))

# What is the average arrival delay for each destination airport?
arr_delay <- flights %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE))

# Find all the flights that flew over 400 miles. What were their destinations?
# Version 1:
long_distance_A <- flights %>%
  filter(distance > 400) %>%
  group_by(dest) %>%
  summarize() %>% # if we don't calculate new columns, we just have the unique destinations 
  left_join(airports, by = c("dest" = "faa")) %>% # to find names
  select(dest, name)

# Version 2:
long_distance_B <- flights %>%
  filter(distance > 400) %>%
  select(dest) %>%
  unique() %>% # removes duplicate entries
  left_join(airports, by = c("dest" = "faa")) %>% # to find names
  select(dest, name)
  
# Which airport was the most common destination? (What is its name?)
common_dest <- flights %>%
  group_by(dest) %>%
  summarize(num_flights = n()) %>%
  arrange(desc(num_flights)) %>%
  head(1) %>%
  left_join(airports, by = c("dest" = "faa")) # find airport information

# Are there any planes in the `flights` table that don't appear in the `planes` table?
missing_planes <- flights %>%
  filter(!is.na(tailnum)) %>% # remove flights with missing tail numbers
  group_by(tailnum) %>%
  summarize() %>% # unique tailnums
  left_join(planes, by = c("tailnum" = "tailnum")) %>%
  filter(is.na(type)) %>%
  nrow() # count how many missing planes

# Among the 100 most commonly used planes, what is the most common manufacturer? Most common model?
# 100 most commonly used planes
common_planes <- flights %>%
  filter(!is.na(tailnum)) %>% # remove flights with missing tail numbers
  group_by(tailnum) %>%
  summarize(num_flights = n()) %>%
  arrange(desc(num_flights)) %>%
  head(100)

# most common manufacturer
common_manufacturer <- common_planes %>%
  left_join(planes, by = c("tailnum" = "tailnum")) %>% # get manufacturer data
  group_by(manufacturer) %>%
  summarize(num_made = n()) %>%
  arrange(desc(num_made)) %>%
  head(1)

# most common model
common_model <- common_planes %>%
  left_join(planes, by = c("tailnum" = "tailnum")) %>% # get manufacturer data
  group_by(model) %>%
  summarize(num_flown = n()) %>%
  arrange(desc(num_flown)) %>%
  head(1)

# untidy data: lengths and albums of Bill Withers songs
untidyA <- tribble(
  ~artist, ~song, ~type, ~info,
  "Bill Withers", "Ain't No Sunshine", "Length", "2:05",
  "Bill Withers", "Ain't No Sunshine", "Album", "Just As I Am",
  "Bill Withers", "Lovely Day", "Length", "4:15",
  "Bill Withers", "Lovely Day", "Album", "Menagerie",
  "Bill Withers", "Lean on Me", "Length", "4:19",
  "Bill Withers", "Lean on Me", "Album", "Still Bill",
  "Bill Withers", "Use Me", "Length", "3:48",
  "Bill Withers", "Use Me", "Album", "Still Bill",
  "Bill Withers", "Grandma's Hands", "Length", "2:01",
  "Bill Withers", "Grandma's Hands", "Album", "Just As I Am"
)

# tidy version: pivot wider (There are multiple rows per observation/song.)
tidyA <- untidyA %>%
  pivot_wider(names_from = "type", values_from = "info")

# untidy data: most popular songs on Spotify, for an eclectic mix of artists from Kathleen's playlists
untidyB <- tribble(
  ~artist, ~`1`, ~`2`, ~`3`, ~`4`, ~`5`,
  "Bill Withers", "Ain't No Sunshine", "Lovely Day", "Lean on Me", "Use Me", "Grandma's Hands",
  "Led Zeppelin", "Stairway to Heaven", "Immigrant Song", "Whole Lotta Love", "Black Dog", "Good Times Bad Times",
  "The Supremes", "You Can't Hurry Love", "Baby Love", "Stop! In The Name Of Love", "Where Did Our Love Go", "Come See About Me",
  "Alessia Cara", "Scars To Your Beautiful", "Out Of Love", "Stay", "Ready", "I Choose",
  "Hozier", "Take Me To Church", "The Bones", "Almost", "Cherry Wine", "Like Real People Do"
)

# tidy version: pivot longer (The column names 1-5 are a variable, rank.)
tidyB <- untidyB %>%
  pivot_longer(c(`1`, `2`, `3`, `4`, `5`), names_to = "rank", values_to = "title")
```
