---
title: "Day 1: Introduction to RStudio & the Tidyverse"
author: "Kathleen Hablutzel"
date: "6/8/2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Welcome!!!

#### Getting Started

If you haven't already, start downloading R and R Studio. See the instructions [here](index.html) under "Getting Started."

Today's complete code is linked [here]().

#### Creating an RMarkdown Document

* Write text, embed code chunks
* R is the language (in console and code chunks)
* RStudio is the programming environment
* RMarkdown is a markup language for combining text with code
    + Produce reproducible reports where the code is right there
    + Export to HTML, PDF, MS Word (how I created website)
    
```{r}
summary(cars)
```

See the RMarkdown [reference guide](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf).

## Data Visualizations with ggplot2

* Make a plot
* Plot elements
* Add colors (colorbrewer)
* Add theme -> point to documentation
* Extension: copy the master examples (a couple graphs for students to copy, then I'll post the code next week and they can see how close they got)

## Visual Cues

* Strongest/weakest cues
* Why stacked bar charts >>>>>>>> pie charts

Ranking visual cues, from position (strongest) to color hue (weakest): docs/day-1_files/figure-html/visual-cues.png (link not  working rn)
Source: Nathan Yau, *Data Points*, 2013