---
title: "Welcome to SRIP Data Science Bootcamp!"
author: "Kathleen Hablutzel"
date: "6/6/2020"
output: 
  html_document:
    fig_width: 9
    fig_height: 2
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  echo = TRUE
)
library(tidyverse)
```

```{r Example_Plot, echo = FALSE} 
housing <- read.csv("http://www.science.smith.edu/~jcrouser/SDS192/landdata-states.csv")

home_plot <- ggplot(data = housing, aes(x = Home.Price.Index, y = State)) +
  geom_point(aes(color = Date),
             alpha = 0.6,
             size = 1.5,
             position = position_jitter(width = 0, height = 0.25)) +
  scale_y_discrete(name = "State Abbreviation") +
  scale_color_continuous(breaks = c(1975.25, 1994.25, 2013.25), # c() list of breaks
                         labels = c(1975, 1994, 2013),
                         low = "blue", high = "red") + # low and high colors
  theme(text = element_blank(),
        rect = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

home_plot
```

## What You'll Learn

#### Skills

- Data Visualization
- Data Wrangling

#### Tools
- R, RStudio, Tidyverse Packages
- GitHub

My goal is to expose you to the first half of a college-level Introduction to Data Science course. We'll meet once a week to code together (tentatively Thursday nights at 8), and then I will provide additional resources if you'd like to delve deeper on each topic.

*Note: You are welcome at all five weeks of SRIP social activities even if your program is shorter than 5 weeks. I will have content for all 5 weeks of SRIP, but I will also provide a reasonable conclusion for those who want to stop after 3 weeks.*

## Getting Started

Please take these steps before our first day:

- [Download and install R](https://cloud.r-project.org/)
- [Download and install RStudio](https://rstudio.com/products/rstudio/download/)[^chrome]
- [Create a GitHub account](https://github.com/)

Now, go into RStudio. In the console at the bottom, install the Tidyverse packages:

```{r Install-Packages, eval = FALSE}
install.packages("tidyverse")
```

If you can run the following code successfully (in the console is fine), you're ready to go!

```{r Example_Code, fig.width = 4, fig.height = 3}
library(tidyverse)
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

If you run into any problems, send me a direct message on Slack and we'll troubleshoot.

[^chrome]: If you have a Chromebook, you'll have to work from an RStudio server. Slack me and we'll get you up and running.